Large Scale Structure Reconstruction with Short-Wavelength Modes

### 2019-10-16 paper/v1.tex
1

\section{Quadratic Estimator}\label{sec2}
In this section we construct the quadratic estimator for long-wavelength modes starting from \ec{2pt}. We then forecast the detectability of the quadratic estimator for future surveys. As with the case of CMB lensing, we can write the general form of the estimator by averaging over pairs of short-wavelength modes:
\begin{eqnarray}
\hat{\delta}^{(1)}(\vec{k}_l)=A(\vec{k}_l)\int \frac{d^3 \vec{k}_s}{(2\pi)^3} g(\vec{k}_s,\vec{k}_s'){\delta}(\vec{k}_s){\delta}(\vec{k}_s') \eql{quadest},
\end{eqnarray} 
with $g$ being a weighting function, $\vk_s'=\vk_l-\vk_s$ and $A$  defined by requiring that $\langle \hat{\delta}^{(1)}(\vec{k}_l) \rangle={\delta}^{(1)}(\vec{k}_l)$:
\begin{eqnarray}
A(\vec{k}_l)=\bigg[\int \frac{d^3 \vec{k}_s}{(2\pi)^3} g(\vec{k}_s,\vec{k}_s')f(\vec{k}_s,\vec{k}_s')  \bigg]^{-1} \eql{a}
\end{eqnarray}
In the absence of shot noise the Gaussian noise is given by:
\be 
\langle \hat{\delta}^{(1)}(\vk_{l})\hat{\delta}^{(1)}(\vk_{l}') \rangle = (2\pi)^3 \delta_{\rm D}(\vk_{l}-\vk_{l}')[P_{\rm lin}(k_{l})+N(\vk_l)]
\ee 
with 
\begin{eqnarray}
&&N(\vec{k}_{l})=2A^2(\vk_{l})\vs
&&\times\int \frac{d^3 \vec{k}_{s}}{(2\pi)^3} g^2(\vec{k}_{s},\vk_l-\vec{k}_{s})P_{\rm nl}(k_{s})P_{\rm nl}(|\vk_l-\vk_s|)
\end{eqnarray}
where $P_{\rm nl}$ is the nonlinear power spectrum. Minimizing the noise term we can fix the form of $g$ to be:
\begin{eqnarray}
g(\vec{k}_{s},\vec{k}_{s}')=\frac{f(\vec{k}_{s},\vec{k}_{s}')}{2P_{\rm nl}(k_{s})P_{\rm nl}(k_{s}')}
\end{eqnarray} 
The noise term reduces simply to $N(\vk_l)=A(\vk_l)$. We find by testing that the value of $N$ is very insensitive to the choice of the lower limit of the integration \ec{a}, since most of the contribution comes from large $k_s$. We discuss the choice of the upper limit later, in section \ref{sec3}.\\
Assuming Gaussian noise, the projected detectability of a $P(k_l)$ measurement using the quadratic estimator can be expressed as:
\be
\frac{1}{\sigma^{2}(k_l)}=\frac{V k_l^2 \Delta k }{(2\pi)^2}\bigg[\frac{P_{\rm lin}(k_l)}{P_{\rm lin}(k_l)+N(k_l) }\bigg]^2 \eql{error},
\ee
where $V$ is the volume of a survey and we compute the detectability for a set of narrow $k_l$-bins each separated by width $\Delta k$.\RC{ Write a couple of sentences about Figure 1, bearing in mind that you have put information in the caption already. Also remember to change the survey volume to match DESI and state this.} \peikai{We show the projected error of long-wavelength power spectrum using this quadratic estimator with fiducial value $1$. The upper limit of the $\vk_s$ integration in \ec{a} is set to be $0.22\,h\,\rm Mpc^{-1}$. We will see in the next section \ref{sec3} that this choice of the upper limit is very reasonable for our current construction at $z=0$. Also notice that $P_{\rm lin}(k_l)$ dominates over $N(k_l)$ in \ec{error} for this upper limit, thus the projected error bars are only slightly wider than the cosmic variance error bars ($N=0$ case).}\\
\peikai{plot linear or nonlinear? Change ylabel?}
\Sfig{SN}{Long-wavelength power spectrum and its error from \ec{error} which can be expressed as $P(k_l)\sigma(k_l)$. We assume a toy survey of boxsize $L=4\, h\rm\, Gpc$, thus volume $V=L^3$ and width $\Delta k = 2\pi/L$. Moreover, we set the integration range for $\vk_s$ from $0.03 \,h \,\rm Mpc^{-1}$ to $0.22\,h\,\rm Mpc^{-1}$.}

\section{Demonstration with an N-Body Simulation}\label{sec3}
We test the power of the quadratic estimator using data from a cosmological N-body simulation. We use the $z=0$ snapshot from BigMPDL, one of the MultiDark cosmological simulations \cite{Klypin:2014nov}. The cubical box side length of BigMDPL is $3.7\,\rm Gpc$. We use the dark matter particle data to compute the matter density field, leaving the effect of using galaxies or halos to trace the field \cite{Desjacques:2018rev} to  future work.\\ 
We use the code nbodykit \cite{Hand:2018nby} to measure the Fourier density modes, and \ec{quadest} to estimate the long wavelength modes  from the measured short wavelength modes. How well the estimator works can be seen from Fig.~\rf{hist}, where we show histograms of the ratio of the estimated mode amplitudes $\hat{\delta}(\vk_l)$ to their true amplitudes $\delta(\vk_l)$. The two panels show the differences between a short wavelength mode cutoff of $k_s=0.22\, h\, \rm Mpc^{-1}$ and  $k_s=0.37\,h\, \rm Mpc^{-1}$. \peikai{Notice that second-order SPT gets less accurate higher in the nonlinear regime and will produce a bias of our quadratic estimator. As we can see in this plot that individual mode amplitudes are basically unbiased when $k_{s,\rm max}= 0.22\,h\, \rm Mpc^{-1}$. While for $k_{s,\rm max}=0.37\,h\, \rm Mpc^{-1}$, the results are more biased (the center of the ratio $\sim 20 \%$ too high) and more incorrect (about $\sim 50 \%$ of the modes have negative real parts of $\delta/\hat{\delta}$) due to the increasing inaccuracy of second-order SPT. } \RC{Comment on what the figure shows -  for example individual mode amplitudes are unbiased when $k_s= 0.15 \rm Mpc^{1}$, and predicted to within $\sim x \%$} \RC{moved this here from fig caption: We can see that the results for $k_s=0.25 \rm Mpc^{1}$ are biased (the mode is $\sim 20 \%$ too high) due to the increasing inaccuracy of second-order SPT as we move into nonlinear scales}
\Sfig{hist}{Comparison of measured and predicted Fourier mode amplitudes: histograms of number counts of $|\hat{\delta}(\vk_l)/\delta(\vk_l)|$ for two different $k_s$ integration ranges.}
\\Another goal of this approach is to reproduce large scale structure in real space. To test this, we compute real space density fields by taking the inverse transform of both the directly measured long wavelength Fourier modes and the long wavelength modes predicted using the quadratic estimator. These density fields, $\delta$ and $\hat{\delta}$ respectively are plotted in Fig.~\rf{real}. We can see that \peikai{our quadratic estimator }
\RC{describe how well it works based on what we can see in the figure}
\Sfigg{real}{Comparison of the true density field in the BigMPDL simulation ($\delta(\vec{x})$ computed using the directly measured large-scale modes, top row) and the density field from the quadratic estimator ($\hat{\delta}(\vec{x})$, middle row). The bottom row shows their difference. Each panel represents a slice through the simulation volume,  $2.5 \,h^{-1}\,\rm Gpc$ wide, and one cell ($0.36 \,h^{-1}\,\rm Gpc$) thick.  The upper limit of $\vk_s$ was $0.22 \,h\, \rm Mpc^{-1}$.}

\section{Conclusion}\label{sec4}
In this paper, we have proposed a new and potentially powerful method to measure long-wavelength modes without having to actually measure large scale structure directly.
\RC{ State that this is good because we won't be sensitive to observational systematics, sky gradients, dust etc}
\RC{Give a short list of cool things we can do, with references: measure 3D clustering beyond the matter radiation equality turnover for the first time, look for evidence of general relativistic effects which are strongest on large scales (e.g., \cite{Jeong:2012ls}), look for evidence of large scale anomalies, like in the CMB, constrain non-Gaussianity,  etc }\\
Extensions of our work suggest themselves and will be relevant to attempts to apply it to observational data. The noise term $N=A$ would reduce significantly if we go higher into the nonlinear regime while second-order SPT is no longer accurate in this case. Thus, to detect even larger scales, we may need to use contributions from higher order perturbations and expand the integration range to reduce the noise level. \\
Another issue relates to the range of long-wavelength modes that are detectable. In the N-body simulation,  $k_l$ for modes recovered ranged from $1.7 \times 10^{-3} \, \rm Mpc^{-1}$ to $8.9\times 10^{-3} \, \rm Mpc^{-1}$.If we try to measure smaller modes, the quadratic estimator approach would become inaccurate since the contraction \ec{contraction} is no longer valid in this case.\\ 
Our current formalism for the quadratic estimator for long-wavelength modes is only valid in a box at a fixed time. As the length scales are so large, a more realistic situation should deal with observations on the light cone. The expression for the quadratic estimator to be generalized to take this into account. Moreover, we notice that the upper limit of integration range should also increase if we consider higher redshifts, in order to include more nonlinear information. Finally, when applied to datasets consisting of halos or galaxies, the approach should be modified to include galaxy bias. We also leave this to future work.\\

>>>>

\section{Quadratic Estimator}\label{sec2}
In this section we construct the quadratic estimator for long-wavelength modes starting from \ec{2pt}. We then forecast the detectability of the quadratic estimator for future surveys. As with the case of CMB lensing, we can write the general form of the estimator by averaging over pairs of short-wavelength modes:
\begin{eqnarray}
\hat{\delta}^{(1)}(\vec{k}_l)=A(\vec{k}_l)\int \frac{d^3 \vec{k}_s}{(2\pi)^3} g(\vec{k}_s,\vec{k}_s'){\delta}(\vec{k}_s){\delta}(\vec{k}_s') \eql{quadest},
\end{eqnarray} 
with $g$ being a weighting function, $\vk_s'=\vk_l-\vk_s$ and $A$  defined by requiring that $\langle \hat{\delta}^{(1)}(\vec{k}_l) \rangle={\delta}^{(1)}(\vec{k}_l)$:
\begin{eqnarray}
A(\vec{k}_l)=\bigg[\int \frac{d^3 \vec{k}_s}{(2\pi)^3} g(\vec{k}_s,\vec{k}_s')f(\vec{k}_s,\vec{k}_s')  \bigg]^{-1} \eql{a}
\end{eqnarray}
In the absence of shot noise the Gaussian noise is given by:
\be 
\langle \hat{\delta}^{(1)}(\vk_{l})\hat{\delta}^{(1)}(\vk_{l}') \rangle = (2\pi)^3 \delta_{\rm D}(\vk_{l}-\vk_{l}')[P_{\rm lin}(k_{l})+N(\vk_l)]
\ee 
with 
\begin{eqnarray}
&&N(\vec{k}_{l})=2A^2(\vk_{l})\vs
&&\times\int \frac{d^3 \vec{k}_{s}}{(2\pi)^3} g^2(\vec{k}_{s},\vk_l-\vec{k}_{s})P_{\rm nl}(k_{s})P_{\rm nl}(|\vk_l-\vk_s|)
\end{eqnarray}
where $P_{\rm nl}$ is the nonlinear power spectrum. Minimizing the noise term we can fix the form of $g$ to be:
\begin{eqnarray}
g(\vec{k}_{s},\vec{k}_{s}')=\frac{f(\vec{k}_{s},\vec{k}_{s}')}{2P_{\rm nl}(k_{s})P_{\rm nl}(k_{s}')}
\end{eqnarray} 
The noise term reduces simply to $N(\vk_l)=A(\vk_l)$. We find by testing that the value of $N$ is very insensitive to the choice of the lower limit of the integration \ec{a}, since most of the contribution comes from large $k_s$. We discuss the choice of the upper limit later, in section \ref{sec3}.\\
Assuming Gaussian noise, the projected detectability of a $P(k_l)$ measurement using the quadratic estimator can be expressed as:
\be
\frac{1}{\sigma^{2}(k_l)}=\frac{V k_l^2 \Delta k }{(2\pi)^2}\bigg[\frac{P_{\rm lin}(k_l)}{P_{\rm lin}(k_l)+N(k_l) }\bigg]^2 \eql{error},
\ee
where $V$ is the volume of a survey and we compute the detectability for a set of narrow $k_l$-bins each separated by width $\Delta k$.\RC{ Write a couple of sentences about Figure 1, bearing in mind that you have put information in the caption already. Also remember to change the survey volume to match DESI and state this.} \peikai{We show the projected error of long-wavelength power spectrum using this quadratic estimator with fiducial value $1$. The upper limit of the $\vk_s$ integration in \ec{a} is set to be $0.22\,h\,\rm Mpc^{-1}$. We will see in the next section \ref{sec3} that this choice of the upper limit is very reasonable for our current construction at $z=0$. Also notice that $P_{\rm lin}(k_l)$ dominates over $N(k_l)$ in \ec{error} for this upper limit, thus the projected error bars are only slightly wider than the cosmic variance error bars ($N=0$ case).}\\
\peikai{plot linear or nonlinear? Change ylabel?}
\Sfig{SN}{Long-wavelength power spectrum and its error from \ec{error} which can be expressed as $P(k_l)\sigma(k_l)$. We assume a toy survey of boxsize $L=4\, h^{-1}\rm\, Gpc$, thus volume $V=L^3$ and width $\Delta k = 2\pi/L$. Moreover, we set the integration range for $\vk_s$ from $0.03 \,h \,\rm Mpc^{-1}$ to $0.22\,h\,\rm Mpc^{-1}$.}

\section{Demonstration with an N-Body Simulation}\label{sec3}
We test the power of the quadratic estimator using data from a cosmological N-body simulation. We use the $z=0$ snapshot from BigMPDL, one of the MultiDark cosmological simulations \cite{Klypin:2014nov}. The cubical box side length of BigMDPL is $2.5\,h^{-1}\,\rm Gpc$. We use the dark matter particle data to compute the matter density field, leaving the effect of using galaxies or halos to trace the field \cite{Desjacques:2018rev} to  future work.\\ 
We use the code nbodykit \cite{Hand:2018nby} to measure the Fourier density modes, and \ec{quadest} to estimate the long wavelength modes  from the measured short wavelength modes. How well the estimator works can be seen from Fig.~\rf{hist}, where we show histograms of the ratio of the estimated mode amplitudes $\hat{\delta}(\vk_l)$ to their true amplitudes $\delta(\vk_l)$ \peikai{for different values of $\vk_l$}. The two panels show the differences between a short wavelength mode cutoff of $k_s=0.22\, h\, \rm Mpc^{-1}$ and  $k_s=0.37\,h\, \rm Mpc^{-1}$. \peikai{Notice that second-order SPT gets less accurate higher in the nonlinear regime and will produce a bias of our quadratic estimator. As we can see in this plot that individual mode amplitudes are basically unbiased when $k_{s,\rm max}= 0.22\,h\, \rm Mpc^{-1}$. While for $k_{s,\rm max}=0.37\,h\, \rm Mpc^{-1}$, the results are more biased (the center of the ratio $\sim 20 \%$ too high) and more incorrect (about $\sim 50 \%$ of the modes have negative real parts of $\delta/\hat{\delta}$) due to the increasing inaccuracy of second-order SPT. } \RC{Comment on what the figure shows -  for example individual mode amplitudes are unbiased when $k_s= 0.15 \rm Mpc^{1}$, and predicted to within $\sim x \%$} \RC{moved this here from fig caption: We can see that the results for $k_s=0.25 \rm Mpc^{1}$ are biased (the mode is $\sim 20 \%$ too high) due to the increasing inaccuracy of second-order SPT as we move into nonlinear scales}
\Sfig{hist}{Comparison of measured and predicted Fourier mode amplitudes: histograms of number counts of $|\hat{\delta}(\vk_l)/\delta(\vk_l)|$ for two different $k_s$ integration ranges.}
\\Another goal of this approach is to reproduce large scale structure in real space. To test this, we compute real space density fields by taking the inverse transform of both the directly measured long wavelength Fourier modes and the long wavelength modes predicted using the quadratic estimator. These density fields, $\delta$ and $\hat{\delta}$ respectively are plotted in Fig.~\rf{real}. We can see that \peikai{our quadratic estimator }
\RC{describe how well it works based on what we can see in the figure}
\Sfigg{real}{Comparison of the true density field in the BigMPDL simulation ($\delta(\vec{x})$ computed using the directly measured large-scale modes, top row) and the density field from the quadratic estimator ($\hat{\delta}(\vec{x})$, middle row). The bottom row shows their difference. Each panel represents a slice through the simulation volume,  $2.5 \,h^{-1}\,\rm Gpc$ wide, and one cell ($0.36 \,h^{-1}\,\rm Gpc$) thick.  The upper limit of $\vk_s$ was $0.22 \,h\, \rm Mpc^{-1}$.}

\section{Conclusion}\label{sec4}
In this paper, we have proposed a new and potentially powerful method to measure long-wavelength modes without having to actually measure large scale structure directly.
\RC{ State that this is good because we won't be sensitive to observational systematics, sky gradients, dust etc}
\RC{Give a short list of cool things we can do, with references: measure 3D clustering beyond the matter radiation equality turnover for the first time, look for evidence of general relativistic effects which are strongest on large scales (e.g., \cite{Jeong:2012ls}), look for evidence of large scale anomalies, like in the CMB, constrain non-Gaussianity,  etc }\\
Extensions of our work suggest themselves and will be relevant to attempts to apply it to observational data. The noise term $N=A$ would reduce significantly if we go higher into the nonlinear regime while second-order SPT is no longer accurate in this case. Thus, to detect even larger scales, we may need to use contributions from higher order perturbations and expand the integration range to reduce the noise level. \\
Another issue relates to the range of long-wavelength modes that are detectable. In the N-body simulation,  $k_l$ for modes recovered ranged from $2.5 \,h \times 10^{-3}  \rm Mpc^{-1}$ to $1.3\,h\times 10^{-2}  \rm Mpc^{-1}$.If we try to measure smaller modes, the quadratic estimator approach would become inaccurate since the contraction \ec{contraction} is no longer valid in this case.\\ 
Our current formalism for the quadratic estimator for long-wavelength modes is only valid in a box at a fixed time. As the length scales are so large, a more realistic situation should deal with observations on the light cone. The expression for the quadratic estimator to be generalized to take this into account. Moreover, we notice that the upper limit of integration range should also increase if we consider higher redshifts, in order to include more nonlinear information. Finally, when applied to datasets consisting of halos or galaxies, the approach should be modified to include galaxy bias. We also leave this to future work.\\
